---
title: "Vowel Analysis Final Report"
author: 
  - "Faith Hardie"
  - "DS303, SP25"
  - "Prof. Amber Camp"
date: 3/14/25
format: html
toc: true
editor: visual
theme: spacelab
---

## Vowel Analysis Final Report

### Load packages

```{r, echo = FALSE, message = FALSE}
library(tidyverse)
library(lme4)
library(lmerTest)

# install.packages("phonR")
library(phonR)
```

## Load data

Load your personal data (make sure you update from P101 -\> your P#)

```{r}
# read in data
P112 <- read_csv("data/P112 (2).csv")

# convert variables to factor where needed
convert_to_factor <- function(df, cols) {
  df[cols] <- lapply(df[cols], as.factor)
  return(df)
}

P112 <- convert_to_factor(P112, c("ppt", "word", "ipa", "arpa", "onset", "offset", "environment", "real_word", "sex", "ethnicity", "birthplace", "home_state", "years_lived_hi", "L1", "exposure_langs_yn", "exposure_langs", "speak_other", "item_num", "rep"))

# remove a couple words you won't be needing
P112 <- P112 %>%
  dplyr::filter(!word %in% c("cot", "caught")) # added dplyr to specify which 'filter' to use

```

Class data:

```{r}
# read in data
all_data <- read_csv("data/DS303_combined.csv")

# convert variables to factor where needed
all_data <- convert_to_factor(all_data, c("ppt", "word", "ipa", "arpa", "onset", "offset", "environment", "real_word", "sex", "ethnicity", "birthplace", "home_state", "years_lived_hi", "L1", "exposure_langs_yn", "exposure_langs", "speak_other", "item_num", "rep"))

# remove a couple words you won't be needing
all_data <- all_data %>%
  dplyr::filter(!word %in% c("cot", "caught"))

```

```{r}
##Summary and skim
#summary(P112)
skimP112 <-  skimr::skim(P112)
#view(P112)

#summary(all_data)
skimalldata <-  skimr::skim(all_data)
view(all_data)
class(all_data)
```

```{r}
##Removing outliers 
## remove outliers
class_clean <- all_data %>%
  group_by(ppt, ipa) %>% 
  mutate(
    f1_z = (f1 - mean(f1)) / sd(f1),
    f2_z = (f2 - mean(f2)) / sd(f2)
  ) %>%
  filter(abs(f1_z) <= 1.25, abs(f2_z) <= 1.25)

P112_clean <- P112 %>%
  group_by(ppt, ipa) %>% 
  mutate(
    f1_z = (f1 - mean(f1)) / sd(f1),
    f2_z = (f2 - mean(f2)) / sd(f2)
  ) %>%
  filter(abs(f1_z) <= 1.25, abs(f2_z) <= 1.25)
```

## Explain the Data

(1 point)

In paragraph form:

-   Describe where the data comes from
-   Summarize the contents of the data (how many observations, participants, items, etc.)
-   Mention any pre-processing steps taken. For example, I pre-processed this data by removing words that were obviously mispronounced before even sending it to you. Then, above, you converted certain variables to factor and removed the words "cot" and "caught", which are not relevant to your investigation. Have you done any additional processing?

**ANSWER:** The data in DS303_combined.csv comes from our classes recorded voices. The process of getting this data started with completing a google form with questions regarding our personal backgrounds such as gender, languages we speak, our place of birth, and other questions based on circumstances that effect how we speak. We then recorded our voices using PRATT software and Dr. Camp organized our classes recordings into a data set.

The data for the class wide data set, DS303_combined.csv has 26 variables, 13 participants, and 1201 items. Each student had 29 words.

In addition to Dr. Camp removing mispronounced words, we converted certain variables to factors and removed the words "cot" and "caught." Independently, I cleaned the data set by removing any outliers by filtering it.

## Variables of Interest

(1 point)

For this project, you will explore and analyze the [**class-wide data set**]{.underline}. In paragraph form:

-   Briefly introduce the purpose of this project
-   Identify and explain your variables of interest
-   State research questions or hypotheses about this data

The purpose of this project is to analyze the different components of how our class says words, including the vowel, onset, offset, and environmental factors that can affect how we speak. My variables of interest are real word, f1, and f2. I am interested to see if whether the word is real or not affects how the vowel is pronounced.

## EDA and Vowel Plots

(3 points)

-   Generate two vowel plots using `phonR`: one for your own data, and one for class data

-   In a couple sentences, state your observations. Do you see any patterns or differences?

-   Include at least one visual that supports your hypothesis/justifies your models below, and explain

```{r}
ggplot(P112, aes(x = real_word, y = f1)) +
  geom_boxplot()
ggplot(class_clean, aes(x=real_word, y=f2)) +
  geom_boxplot()
ggplot(P112, aes(x = real_word, y = f1)) +
  geom_boxplot()
ggplot(class_clean, aes(x=real_word, y=f2)) +
  geom_boxplot()
#ggplot(class_clean, aes(x= f1, y=f2, fill=ethnicity))+
#  geom_point()
par()
with(class_clean, plotVowels(f1, f2, var.sty.by=sex, var.col.by = real_word, pretty = TRUE,  legend.kwd = "bottomleft"))
```

**Patterns, observations:** For the EDA part of this assignment, I did 4 box plots using ggplot and 1 dot plot using phonr. For both my personal data and the class data, there is a small difference in both formant 1 and 2 depending on whether the word is real or not real, however, this doesn't necessary mean that the vowel formants changed because of the words realness, it could be dependent on other variables, how how the words that aren't real are commonly pronounced (the way the tongue sits when the words are said) compared to if the word is real. The difference between real and not real also doesn't seem to be statistically significant from these graphs, but I would have to run my regression models to fully understand the relationship.

## Model Selection and Justification

(3 points)

-   You will build and analyze **two different statistical models** to investigate the relationship between your predictors and outcome variable

-   The two models should differ in some way (e.g., one could include an additional or interaction term while the other does not)

-   What statistical models will you use to investigate the relationship between your predictors and outcome variable? (linear vs. logistic regression? mixed effects model?)

-   Why did you select these models?

-   Which variable(s) are included?

    Answer: I will be using logistic regression to investigate the relationship between my predictors and outcome variable because my predictor variable real_word has only two options, real or not_real, so it is a binomial predictor. However, I also wanted to investigate if f1 and f2 could predict if the word is real or not, and for that I need to use linear regression, since the predictor f1 is not a binary or binomial predictor, and its values are outside of 0 or 1. The variables I am including are real_word, f1, f2, and sex.

```{r}
m1 <- glm( real_word ~ f1 , data = class_clean, family = binomial)
summary(m1)
m2 <- glm(real_word ~ f2, data = class_clean, family = binomial) 
summary(m2) 
#m3 <- glm(real_word ~ f1*sex, data = class_clean, family = binomial) 
#summary(m3) 
m4 <- glm(real_word ~ f2*sex, data = class_clean, family = binomial) 
summary(m4) 
```

```{r}
m5 <- lm(f1 ~ real_word, data=class_clean) 
summary(m5) 
class_clean$sex <- relevel(class_clean$sex, ref = "Male") 
m6 <- lm(f1 ~ real_word *sex, data=class_clean) 
summary(m6) 
```

## Model Comparisons and Best Fit

(3 points)

-   Build and run both models and display their summaries

-   Compare the two models, assess model fit, and determine the better fitting one

```{r}
anova(m1, m3)
#anova(m2, m4)
#anova(m5, m6)
```

## Interpretation of Results

(3 points)

-   Interpret coefficients and significance
-   Explain how the predictor variable(s) influence the outcome

**Answer:** Model 1 and 3:

## Discussion and Conclusion

(3 points)

-   Summarize key findings
-   Discuss implications
-   Mention limitations
